### 2020-06-26 paper/sex_selection_spacing-ver4.tex
Edit for clarity and conciseness
Down to 879 words, not including references

% Background
India saw many positive changes over the last four decades:
the economy grew substantially,
schooling increased for both males and females,
and the total fertility rate fell to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

However, India also witnessed the appearance of sex selection---the selective 
abortion of female fetuses based on prenatal sex determination. 
Sex selection became available in the mid-eighties with the introduction of technologies 
such as ultrasound and amniocentesis. 
Combined with continued strong son preference, the result was a dramatic increase in the 
male-to-female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question I examine is: how did birth spacing respond to these
changes, especially the introduction and spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover; 
otherwise, the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is between one and six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation are all also likely to impact birth spacing. 
Hence, the combined changes in birth spacing in India may be much more substantial than
 what we have observed in other countries.

Studying birth spacing is valuable because it contributes to our knowledge about 
fertility decisions and because of birth spacing's potential effects on outcomes such as 
fertility measures and mortality. 
To understand the effects of changes in birth spacing, I focus on two questions.

First, to what extent do changes in birth spacing bias our standard fertility estimates?
Tempo effects from longer birth intervals make the total fertility rate a downward biased 
estimate of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's chances of survival?
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing then potentially lowers mortality and morbidity risk 
\citep{Conde-Agudelo2012,Molitoris2019}.
There may, however, be counteracting effects if the longer spacing arises because of 
multiple abortions.
There has, so far, been no attempt to examine how infant mortality has responded 
to the increase in sex selection.

% How the study progresses
To investigate how birth spacing has changed over time and across groups with the 
introduction of sex selection, I use a competing risk hazard model with two exit 
states: the birth of a girl or the birth of a boy. 
I apply the model to the birth histories of Hindu women covering 1972 to 2016 using 
data from the four National Family and Health Surveys (NFHS).

The primary outcomes I examine are average birth intervals, the sex ratio of births, 
and the likelihood of a next birth. 
Because sex selection might affect the distribution of birth intervals, I also
examine the 25th and 75th percentiles of birth intervals. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key explanatory variables are maternal education, the sex composition of previous 
children, and area of residence.

The empirical model allows me to predict cohort fertility. 
To examine to what extent tempo effects bias our standard fertility measures, I compare 
this predicted cohort fertility with fertility calculated from age-specific fertility rates.

I use the same data to study how infant mortality risk has changed with the changing 
birth spacing and whether sex selection directly affects infant mortality. 
The explanatory variables are the same as for the spacing analysis, except for the 
addition of birth spacing and index child sex.

@@

First, birth intervals have increased for all education groups over the four decades, but 
more so, the higher the parity and the higher the education. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with women 
with no sons having \emph{longer} birth intervals than women with sons. 
Those least likely to use sex selection---women with less education in rural areas---still 
follow the traditional pattern of short spacing when they have girls and little evidence 
of sex selection.

Second, there is little change in the likelihood of having a very short birth interval. 
Hence, the increased average spacing comes mainly from the longer birth intervals getting 
even longer, primarily driven by substantial use of sex selection. 
The exception to this result is that those better-educated women who have a very high 
predicted use of sex selection conceive later precisely because the next child is almost 
certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s when spacing began to increase.
Although predicted cohort fertility and the fertility rate have lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level for all but the best-educated urban women.

Finally, infant mortality risk has declined substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
However, the mortality risk is still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.

>>>>

% Background
India experienced many positive changes over the last four decades:
the economy grew substantially,
schooling increased for both males and females,
and the total fertility rate fell to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-eighties.
Combined with continued strong son preference, the result was a dramatic increase in the 
male-to-female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question here is: how did birth spacing respond to these changes, especially 
the spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation are all likely to impact birth spacing. 
The combined changes in birth spacing may, therefore, outpace what we have observed in 
other countries.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing affects the reliability of our fertility measures and may 
impact mortality. 
To understand these effects, I address two additional questions.

First, did birth spacing changes bias the fertility estimates for India?
The tempo effect from longer birth intervals makes the total fertility rate a downward 
biased estimate of cohort fertility. \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's chances of survival?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing may, therefore, lower mortality 
\citep{Conde-Agudelo2012,Molitoris2019}.
A counteracting effect is, however, possible if the longer spacing arises because of 
multiple abortions. 
Nevertheless, there has been no attempt to examine how infant mortality changed with 
the increase in sex selection.

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: the birth of a girl or the birth of a boy. 
I apply the model to the birth histories of Hindu women covering 1972 to 2016 using 
data from the four National Family and Health Surveys (NFHS).

The primary outcomes I examine are average birth intervals, the sex ratio of births, 
and the likelihood of a next birth. 
Because sex selection might also affect the distribution of birth intervals, I furthermore
examine the 25th and 75th percentiles of birth intervals. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key variables are maternal education, the sex composition of previous 
children, and area of residence.

The empirical model allows me to predict cohort fertility. 
To examine whether tempo effects bias our standard fertility measures, I compare 
the predicted cohort fertility with fertility calculated from age-specific fertility 
rates.

I use the same data to study how infant mortality changed with birth spacing and 
the increasing use of sex selection. 
The key explanatory variables remain the same, except for the addition of birth spacing 
and the sex of the index child.

@@

First, over the four decades, average birth intervals lengthen for all, but most for higher 
parities and more educated women. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated in rural areas---still follow 
the traditional pattern of short spacing when they have girls and little evidence of sex 
selection.

Second, the likelihood of a very short birth interval changed little. 
Hence, the increased average spacing comes mainly from longer birth intervals getting even 
longer, primarily because of sex selection. 
The exception is that those better-educated women with a very high predicted use of sex 
selection conceive later and are less likely to have a very short birth interval, precisely 
because their next child is almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still ten to 
twenty percent higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

Finally, infant mortality has declined substantially over time for all groups, but fastest 
for the less educated, who are now close to the level of the best-educated women. 
However, mortality is still inversely related to education level, especially for very short 
birth intervals. 
There is no evidence that the increasing use of sex selection is associated with higher mortality.
